---
author: Jack Bienvenue
date: 12 November 2025
output: pdf_document
---

### Generalized Pareto Distribution Fit

GPD has three parameters: $\mu$--location, $\sigma$--scale, and $\xi$--shape.

We can estimate the parameters through maximum likelihood estimation.

A few people have built methods for MLE of $\sigma$ and $\xi$. The one used here comes from <https://search.r-project.org/CRAN/refmans/evmix/html/fgpd.html>.

```{r}
## Import packages
library(evmix)
library(dplyr)
library(tidyr)
library(ggplot2)
library(extRemes)

# Import data
asheville_df <- read.csv("../data/ghcn_roi_daily_data/Asheville.csv")
## Remove NAs
asheville_df <- asheville_df %>%
  drop_na()

# GPD Fitting mechanism
set.seed(5845)
fit <- fgpd(asheville_df$precipitation, u = 44.7) # 44.7 mm threshold 
                                # represents 99th percentile of data

# Inspect results
fit$sigmau      # MLE of sigma
fit$xi          # MLE of xi
fit$se          # standard errors
fit$cov         # variance-covariance matrix
fit$nllh        # minimized negative log-likelihood
fit$rate        # estimated tail fraction phi_u (proportion of exceedances)
fit$optim$convergence
```

### NOTE THE STANDARD ERROR OF THE $\xi$ ESTIMATE! A CONFIDENCE INTERVAL WOULD SPAN OVER $\xi=$, SO WE CANNOT CLAIM THAT THE EXCEEDANCES OVER 44.7 mm ARE NOT EXPONENTIALLY-DISTRIBUTED!

## Plot fitted distribution against true points:

```{r}
# Extract threshold and fitted parameters
u <- 44.7
sigma_hat <- fit$sigmau
xi_hat <- fit$xi

# Extract tail data (exceedances)
exceedances <- asheville_df$precipitation[asheville_df$precipitation > u]
y <- exceedances - u  # GPD models the excess over threshold

# Compute fitted GPD density and empirical histogram
# We’ll make a histogram of the exceedances and overlay the fitted pdf
hist_df <- data.frame(y = y)

# Create a sequence of excess values for smooth density curve
y_seq <- seq(0, max(y), length.out = 200)
gpd_pdf <- dgpd(y_seq, sigma = sigma_hat, xi = xi_hat)

# Plot: histogram of empirical excesses + fitted GPD pdf
# Create data frame for the fitted GPD curve
gpd_df <- data.frame(y_seq = y_seq, gpd_pdf = gpd_pdf)

# Plot: histogram of empirical excesses + fitted GPD pdf
ggplot(hist_df, aes(x = y)) +
  geom_histogram(aes(y = ..density..), breaks = seq(0, max(y), length.out = 21),
                 fill = "skyblue", color = "white", alpha = 0.6) +
  geom_line(data = gpd_df, aes(x = y_seq, y = gpd_pdf),
            color = "red", size = 1.1) +
  labs(
    title = "Empirical Tail (Excesses) vs Fitted GPD",
    x = "Excess over threshold (mm)",
    y = "Density"
  ) +
  theme_minimal()
```

```{r return-level-plot, fig.width = 8, fig.height=5.25}
# --- PARAMETERS ---
u <- 44.7
sigma_hat <- 18.5
xi_hat <- 0

# Number of observations and exceedances
N <- nrow(asheville_df)
N_exc <- sum(asheville_df$precipitation > u)

# Tail fraction (rate of exceedances)
lambda <- N_exc / N

# Desired return periods (years)
# Assume daily data -> convert years to days (365.25 days per year)
return_periods_years <- c(5, 10, 50, 100, 500, 1000)
return_periods_days <- return_periods_years * 365.25

# --- CALCULATE RETURN LEVELS ---
if(xi_hat == 0){
  # Exponential GPD formula for return level
  return_levels <- u + sigma_hat * log(lambda * return_periods_days)
} else {
  # General formula for xi != 0
  return_levels <- u + (sigma_hat / xi_hat) * ((lambda * return_periods_days)^xi_hat - 1)
}

# Combine results into a table
results <- data.frame(
  ReturnPeriod_years = return_periods_years,
  ReturnLevel_mm = return_levels
)

print(results)

# --- Extract sigma standard error from fit ---
# This assumes your fgpd fit object provides standard errors
SE_sigma <- fit$se[1]  # adjust name depending on your fit object

# Return periods in days (daily data)
return_periods_days <- return_periods_years * 365.25

# Compute return levels
return_levels <- u + sigma_hat * log(lambda * return_periods_days)

# Compute standard errors of return levels (delta method)
SE_return_levels <- log(lambda * return_periods_days) * SE_sigma

# Compute 95% CI bounds
lower_CI <- return_levels - 1.96 * SE_return_levels
upper_CI <- return_levels + 1.96 * SE_return_levels

# Add to results data frame
results$Lower_CI <- lower_CI
results$Upper_CI <- upper_CI


# -------- GET EMPIRICAL RETURN POINTS

precip_sorted <- sort(asheville_df$precipitation, decreasing = TRUE)

N <- length(precip_sorted)

# Empirical return periods (Weibull plotting formula)
empirical_RP_days <- (N + 1) / seq_along(precip_sorted)  # i = 1,2,...,N
empirical_RP_years <- empirical_RP_days / 365.25

# Combine into a data frame
empirical_df <- data.frame(
  ReturnPeriod_years = empirical_RP_years,
  ReturnLevel_mm = precip_sorted
)

#-------- GET A SPECIAL LABEL FOR HURRICANE HELENE'S WORST DAY:
# Special return level
special_x <- 146.8

# Compute corresponding return period (days)
special_T_days <- (1 / lambda) * exp((special_x - u) / sigma_hat)

# Convert to years
special_T_years <- special_T_days / 365.25

# Create a data frame for the special point
special_point <- data.frame(
  ReturnPeriod_years = special_T_years,
  ReturnLevel_mm = special_x
)

#---------------------#
ggplot() +
  # CI ribbon
  geom_ribbon(data = results, aes(x = ReturnPeriod_years, ymin = Lower_CI, ymax = Upper_CI),
              fill = "lightblue", alpha = 0.3) +
  # Fitted GPD line
  geom_line(data = results, aes(x = ReturnPeriod_years, y = ReturnLevel_mm),
            color = "cyan4", size = 1) +
  # Empirical points
  geom_point(data = empirical_df, aes(x = ReturnPeriod_years, y = ReturnLevel_mm),
             color = "red3", size = 2, alpha = 0.7) +
  # Special labeled point
  geom_point(data = special_point, aes(x = ReturnPeriod_years, y = ReturnLevel_mm),
             color = "darkgreen", size = 4) +
  geom_text(data = special_point,
            aes(x = ReturnPeriod_years, y = ReturnLevel_mm,
                label = paste0("(", round(ReturnPeriod_years,1), " yr, ", ReturnLevel_mm, " mm)")),
            vjust = -1.75, hjust = 0.5, color = "darkgreen", fontface = "bold",
            size = 4) +
  # Log scale with limits
  scale_x_log10(
    limits = c(5, max(c(results$ReturnPeriod_years, empirical_df$ReturnPeriod_years))),
    breaks = c(5, 10, 50, 100, 500, 1000),
    labels = c(5, 10, 50, 100, 500, 1000)
  ) +
  labs(
    title = "Return Level vs Return Period (Empirical + GPD Fit)",
    subtitle = "Empirical return periods vs GPD fit with 95% CI",
    x = "Return Period (years, log scale)",
    y = "Return Level (mm)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```
