---
author: Jack Bienvenue
date: 12 November 2025
output: pdf_document
---

### Generalized Pareto Distribution Fit

GPD has three parameters: $\mu$--location, $\sigma$--scale, and $\xi$--shape. 

We can estimate the parameters through maximum likelihood estimation. 

A few people have built methods for MLE of $\sigma$ and $\xi$. The one used here comes from https://search.r-project.org/CRAN/refmans/evmix/html/fgpd.html.

```{r}
## Import packages
library(evmix)
library(dplyr)
library(tidyr)
library(ggplot2)
library(extRemes)

# Import data
asheville_df <- read.csv("../data/ghcn_roi_daily_data/Asheville.csv")
## Remove NAs
asheville_df <- asheville_df %>%
  drop_na()

# GPD Fitting mechanism
set.seed(5845)
fit <- fgpd(asheville_df$precipitation, u = 44.7) # 44.7 mm threshold 
                                # represents 99th percentile of data

# Inspect results
fit$sigmau      # MLE of sigma
fit$xi          # MLE of xi
fit$se          # standard errors
fit$cov         # variance-covariance matrix
fit$nllh        # minimized negative log-likelihood
fit$rate        # estimated tail fraction phi_u (proportion of exceedances)
fit$optim$convergence
```

### NOTE THE STANDARD ERROR OF THE $\xi$ ESTIMATE! A CONFIDENCE INTERVAL WOULD SPAN OVER $\xi=$, SO WE CANNOT CLAIM THAT THE EXCEEDANCES OVER 44.7 mm ARE NOT EXPONENTIALLY-DISTRIBUTED!

## Plot fitted distribution against true points:

```{r}
# Extract threshold and fitted parameters
u <- 44.7
sigma_hat <- fit$sigmau
xi_hat <- fit$xi

# Extract tail data (exceedances)
exceedances <- asheville_df$precipitation[asheville_df$precipitation > u]
y <- exceedances - u  # GPD models the excess over threshold

# Compute fitted GPD density and empirical histogram
# We’ll make a histogram of the exceedances and overlay the fitted pdf
hist_df <- data.frame(y = y)

# Create a sequence of excess values for smooth density curve
y_seq <- seq(0, max(y), length.out = 200)
gpd_pdf <- dgpd(y_seq, sigma = sigma_hat, xi = xi_hat)

# Plot: histogram of empirical excesses + fitted GPD pdf
# Create data frame for the fitted GPD curve
gpd_df <- data.frame(y_seq = y_seq, gpd_pdf = gpd_pdf)

# Plot: histogram of empirical excesses + fitted GPD pdf
ggplot(hist_df, aes(x = y)) +
  geom_histogram(aes(y = ..density..), breaks = seq(0, max(y), length.out = 21),
                 fill = "skyblue", color = "white", alpha = 0.6) +
  geom_line(data = gpd_df, aes(x = y_seq, y = gpd_pdf),
            color = "red", size = 1.1) +
  labs(
    title = "Empirical Tail (Excesses) vs Fitted GPD",
    x = "Excess over threshold (mm)",
    y = "Density"
  ) +
  theme_minimal()
```

``` {r return-level-plot, fig.height = 8}
library(extRemes)
library(dplyr)
library(ggplot2)

# Load + clean
ashe <- read.csv("../data/ghcn_roi_daily_data/Asheville.csv") %>% drop_na()

# Threshold (99th percentile)
u <- quantile(ashe$precipitation, 0.99)

# Data for POT fit
ashe_pot <- ashe %>% filter(precipitation > u)

# Fit GPD with extRemes
fit_ex <- fevd(
  precipitation,
  threshold = u,
  type = "GP",
  data = ashe,
  method = "MLE"
)

fit_ex


png(filename = "../figures/gpd_subplots.png", width = 800, height = 800)
plot(fit_ex)
dev.off()
plot(fit_ex)

ci(fit_ex)
ci(fit_ex, type = "parameter")

plot(fit_ex, type = "rl")

# Get explicit return levels for 5,10,50,100,500, and 1000-year storms
return.level(fit_ex, return.period = c(5, 10, 50, 100, 500, 1000))

# NOTE: Helene also brought 103.9mm (9/25/24) and 104.4 (9/27/24)
```