---
author: Jack Bienvenue
date: 12 November 2025
output: pdf_document
---

### Generalized Pareto Distribution Fit

GPD has three parameters: $\mu$--location, $\sigma$--scale, and $\xi$--shape. 

We can estimate the parameters through maximum likelihood estimation. 

A few people have built methods for MLE of $\sigma$ and $\xi$. The one used here comes from https://search.r-project.org/CRAN/refmans/evmix/html/fgpd.html.

```{r}
## Import packages
library(evmix)
library(dplyr)
library(tidyr)
library(ggplot2)

# Import data
asheville_df <- read.csv("../data/ghcn_roi_daily_data/Asheville.csv")
## Remove NAs
asheville_df <- asheville_df %>%
  drop_na()

# GPD Fitting mechanism
set.seed(5845)
fit <- fgpd(asheville_df$precipitation, u = 44.7) # 44.7 mm threshold 
                                # represents 99th percentile of data

# Inspect results
fit$sigmau      # MLE of sigma
fit$xi          # MLE of xi
fit$se          # standard errors
fit$cov         # variance-covariance matrix
fit$nllh        # minimized negative log-likelihood
fit$rate        # estimated tail fraction phi_u (proportion of exceedances)
fit$optim$convergence
```

### NOTE THE STANDARD ERROR OF THE $\xi$ ESTIMATE! A CONFIDENCE INTERVAL WOULD SPAN OVER $\xi=$, SO WE CANNOT CLAIM THAT THE EXCEEDANCES OVER 44.7 mm ARE NOT EXPONENTIALLY-DISTRIBUTED!

## Plot fitted distribution against true points:

```{r}
# Extract threshold and fitted parameters
u <- 44.7
sigma_hat <- fit$sigmau
xi_hat <- fit$xi

# Extract tail data (exceedances)
exceedances <- asheville_df$precipitation[asheville_df$precipitation > u]
y <- exceedances - u  # GPD models the excess over threshold

# Compute fitted GPD density and empirical histogram
# We’ll make a histogram of the exceedances and overlay the fitted pdf
hist_df <- data.frame(y = y)

# Create a sequence of excess values for smooth density curve
y_seq <- seq(0, max(y), length.out = 200)
gpd_pdf <- dgpd(y_seq, sigma = sigma_hat, xi = xi_hat)

# Plot: histogram of empirical excesses + fitted GPD pdf
# Create data frame for the fitted GPD curve
gpd_df <- data.frame(y_seq = y_seq, gpd_pdf = gpd_pdf)

# Plot: histogram of empirical excesses + fitted GPD pdf
ggplot(hist_df, aes(x = y)) +
  geom_histogram(aes(y = ..density..), bins = 25,
                 fill = "skyblue", color = "white", alpha = 0.6) +
  geom_line(data = gpd_df, aes(x = y_seq, y = gpd_pdf),
            color = "red", size = 1.1) +
  labs(
    title = "Empirical Tail (Excesses) vs Fitted GPD",
    x = "Excess over threshold (mm)",
    y = "Density"
  ) +
  theme_minimal()
```

``` {r return-level-plot}
# Get other parameter estimates required for return level computation:
phi_hat <- fit$rate # proportion of exceedances
cov_mat <- fit$cov # 2x2 covariance matrix for (sigma, xi)

##### Return levels for a grid of return periods #####

return_periods <- seq(1.01, 500, length.out = 300)  # avoid T=1

# Return level function
zT <- function(T, sigma, xi, phi) {
  u + (sigma/xi) * ((T * phi)^xi - 1)
}

# Compute curve
rl <- zT(return_periods, sigma_hat, xi_hat, phi_hat)

##### Delta-method CI #####

# Gradient of z(T) w.r.t (sigma, xi)
grad_z <- function(T, sigma, xi, phi) {
  A <- T * phi
  d_sigma <- (1/xi) * (A^xi - 1)
  d_xi <- -(sigma/xi^2) * (A^xi - 1) +
           (sigma/xi) * (A^xi * log(A))
  c(d_sigma, d_xi)
}

# Standard errors
se_rl <- sapply(return_periods, function(T) {
  g <- grad_z(T, sigma_hat, xi_hat, phi_hat)
  sqrt(t(g) %*% cov_mat %*% g)
})

# 95% CI
rl_lower <- rl - 1.96 * se_rl
rl_upper <- rl + 1.96 * se_rl

return_df <- data.frame(
  T = return_periods,
  rl = rl,
  lower = rl_lower,
  upper = rl_upper
)

##### Observed exceedances plotted with empirical return periods #####

exceedances <- asheville_df$precipitation[asheville_df$precipitation > u]

# Empirical return period: T_i = 1 / (rank/N exceedance rate)
N_ex <- length(exceedances)
empirical_T <- 1 / ((1:N_ex)/N_ex * phi_hat)  # increasing T

obs_df <- data.frame(
  T = empirical_T,
  z = exceedances
)
View(obs_df)
##### Plot #####

ggplot(return_df, aes(x = T, y = rl)) +
  geom_line(color = "blue", linewidth = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.25, fill = "blue") +
  geom_point(data = obs_df, aes(x = T, y = z), color = "red", alpha = 0.6) +
  scale_x_log10() +
  labs(
    title = "Generalized Pareto Return Level Plot (POT method)",
    x = "Return period T (years, log scale)",
    y = "Return level (mm)"
  ) +
  theme_minimal(base_size = 14)
```