---
title: STAT5845 HW8
author: Jack Bienvenue
date: 17 November 2025
output:
  bookdown::pdf_document2:
    latex_engine: pdflatex
    keep_tex: true 
header-includes:
  - \usepackage{titling}
  - \setlength{\droptitle}{1em} 
fontsize: 12pt
---

```{r echo=FALSE}
# Initialization

## Package Import
invisible(lapply(c("dplyr", "tidyr", "ggplot2",
                   "gstat", "tidyverse", "patchwork",
                   "broom", "lubridate", "fields",
                   "sp", "sf", "rnaturalearth",
                   "rnaturalearthdata", "scales",
                   "grid"), 
  function(pkg) suppressPackageStartupMessages(library(pkg, 
                                                character.only = TRUE))))

## Data Import
data(wind)

## Data cleaning/prep
# Convert to tibble for easier handling
wind_tbl <- as_tibble(wind) %>%
  mutate(date = seq(as.Date("1961-01-01"), as.Date("1978-12-31"), by = "day")) %>%
  relocate(date)

# Long format
wind_long <- wind_tbl %>%
  pivot_longer(cols = RPT:MAL, names_to = "station", values_to = "wind")

# Collapse to monthly averages
wind_month <- wind_long %>%
  mutate(ym = floor_date(date, "month")) %>%
  group_by(station, ym) %>%
  summarise(wind = mean(wind, na.rm = TRUE), .groups = "drop") %>%
  mutate(month = month(ym, label = TRUE, abbr = TRUE))
```

# Gneiting-Matérn Space-time Covariance Model of Residuals via MLE

```{r}
# ======================================================================
# DETERMINISTIC — Build seasonal regressors & fit three models
# ======================================================================

wind_model <- wind_month %>%
  mutate(
    year_num = year(ym) + (month(ym) - 0.5)/12,
    t        = year_num - min(year_num),
    s1 = sin(2*pi*t),
    c1 = cos(2*pi*t)
  )

m_pooled     <- lm(wind ~ s1 + c1, data = wind_model)
m_intercepts <- lm(wind ~ station + s1 + c1, data = wind_model)
m_full       <- lm(wind ~ station * (s1 + c1), data = wind_model)

# --- Extract residuals from best deterministic model (Model C) ----------------
resid_C <- augment(m_full, data = wind_model) %>%
  transmute(station, ym, resid = .resid)

# ============================================================
# Fit Matérn space–time covariance to residuals
# ============================================================
wind.loc$y <- as.numeric(char2dms(as.character(wind.loc$Latitude)))
wind.loc$x <- as.numeric(char2dms(as.character(wind.loc$Longitude)))
coords <- wind.loc %>%
  mutate(
    y = as.numeric(char2dms(as.character(Latitude))),
    x = as.numeric(char2dms(as.character(Longitude)))
  ) %>%
  select(station = Code, x, y)

st_resid <- resid_C %>%
  left_join(coords, by = "station")

# ============================================================
# TRAIN–TEST SPLIT: remove last 6 months
# ============================================================

last6_months <- sort(unique(st_resid$ym), decreasing = TRUE)[1:6]
st_train <- st_resid %>% filter(!ym %in% last6_months)
st_test  <- st_resid %>% filter(ym %in% last6_months)

# ============================================================
# Fit Matérn space–time covariance to Irish monthly residuals
# ============================================================

# Use training residuals as df
df <- st_train %>% arrange(ym, station)

# Response
y <- df$resid
N <- length(y)

# Spatial distance matrix (lon/lat -> great-circle distance)
locs <- as.matrix(df[, c("x", "y")])
D_spatial <- rdist.earth(locs, miles = TRUE)

# Time in months since first month
time_vec <- as.numeric(difftime(df$ym, min(df$ym), units = "days")) / 30.4375
U_time <- abs(outer(time_vec, time_vec, "-"))

# ------------------------------------------------------------
# space–time Matérn covariance function
# ------------------------------------------------------------
matern_corr <- function(D, nu, range_s) {
  if (is.infinite(nu)) return(exp(-(D / range_s)^2))
  
  r <- sqrt(2 * nu) * D / range_s
  C <- (2^(1 - nu) / gamma(nu)) * (r^nu) * besselK(r, nu)
  C[r == 0] <- 1
  C
}

build_K_spacetime <- function(sigma2, beta_s, nu, nug, beta_t, gamma_st,
                              D_spatial, U_time) {
  alpha <- 1
  tau   <- 1   # d/2 for d = 2
  
  S         <- 1 + (U_time^(2 * alpha)) / beta_t      # ψ(|u|^2)
  range_mat <- beta_s * S^(gamma_st / 2)              # ℓ_ij = β_s ψ^{γ/2}
  R_mat     <- matern_corr(D_spatial, nu, range_mat)  # spatial Matérn
  K         <- sigma2 * S^(-tau) * R_mat              # σ² ψ^{-d/2} Matérn
  
  diag(K) <- diag(K) + nug
  K
}

# ------------------------------------------------------------
# Negative log-likelihood and optimization
# ------------------------------------------------------------
nll <- function(par) {
  nug      <- 1e-6
  sigma2   <- par[1]^2
  beta_s   <- exp(par[2])
  beta_t   <- exp(par[3])
  gamma_st <- 1 / (1 + exp(-par[4]))  # logistic transform in (0,1)
  nu_val   <- exp(par[5])
  mu_val   <- 0                       # zero-mean residuals
  
  K <- build_K_spacetime(sigma2, beta_s, nu_val, nug, beta_t, gamma_st,
                         D_spatial, U_time)
  
  L <- tryCatch(chol(K), error = function(e) NULL)
  if (is.null(L)) return(1e10)
  
  z <- backsolve(L, y - mu_val, transpose = TRUE)
  logdetK <- 2 * sum(log(diag(L)))
  quad    <- sum(z^2)
  
  0.5 * N * log(2 * pi) + 0.5 * logdetK + 0.5 * quad
}

# Rough initial values 
init <- c(
  sqrt(var(y, na.rm = TRUE)),                            # sigma2 scale
  log(median(D_spatial[D_spatial > 0], na.rm = TRUE)),   # beta_s
  log(5),                                                # beta_t (months)
  0,                                                     # gamma_st on logit scale
  log(0.5)
)

# To only run the optimization once, I suspended the call:

# fit <- optim(init, nll, control = list(trace = 3, maxit = 500))
# 
# est_par     <- fit$par
# tausq_hat   <- 1e-6
# sigma2_hat  <- est_par[1]^2
# beta_s_hat  <- exp(est_par[2])
# beta_t_hat  <- exp(est_par[3])
# gamma_hat   <- 1 / (1 + exp(-est_par[4]))
# nu_hat      <- exp(est_par[5])
# mu_hat      <- 0
# 
# results <- data.frame(
#   Parameter = c("tau^2 (nugget)", "sigma^2 (variance)", "beta_s (spatial range, miles)",
#                 "beta_t (temporal range, months)", "gamma (space-time)", 
#                 "nu (smoothness)", "mu (mean)"),
#   Estimated = c(tausq_hat, sigma2_hat, beta_s_hat, beta_t_hat,
#                 gamma_hat, nu_hat, mu_hat)
# )

# Instead of re-running that code, just read saved csv:
#------<-------COMMENT OUT THIS SECTION if running optim -------->------#
results <- read.csv("optimization_results.csv")

tausq_hat   <- results$Estimated[1]
sigma2_hat  <- results$Estimated[2]
beta_s_hat  <- results$Estimated[3]
beta_t_hat  <- results$Estimated[4]
gamma_hat   <- results$Estimated[5]
nu_hat      <- results$Estimated[6]
mu_hat      <- results$Estimated[7]

cat("Estimated Parameters (Irish monthly winds):\n")
print(results, digits = 4, row.names = FALSE)

# --------------------------------------------------
# Plot estimated space–time covariance
# --------------------------------------------------

# scalar C(h,u) consistent with build_K_spacetime
matern_spacetime_cov <- function(h, u, sigma2, beta_s, beta_t, gamma_st, nu, nug = 0) {
  alpha <- 1
  tau   <- 1
  
  S_u      <- 1 + (abs(u)^(2 * alpha)) / beta_t   # ψ(|u|^2)
  range_u  <- beta_s * S_u^(gamma_st / 2)         # ℓ(u) = β_s ψ^{γ/2}
  R        <- matern_corr(h, nu, range_u)         # Matérn(h; ℓ(u))
  cov_val  <- sigma2 * S_u^(-tau) * R             # σ² ψ^{-d/2} Matérn
  
  if (nug > 0 && u == 0) cov_val[h == 0] <- cov_val[h == 0] + nug
  cov_val
}

max_d <- max(D_spatial, na.rm = TRUE)
spatial_distance <- seq(0, max_d, length.out = 200)

time_points <- c(0, 1, 3, 6, 12)

colors <- c("#000000", "#004369", "#BE2D2D", "#0e6900", "#FF8C00")

plot_df <- expand.grid(
  h = seq(0, max(D_spatial, na.rm = TRUE), length.out = 200),
  u = time_points
)

plot_df$cov <- mapply(
  FUN = function(h, u) {
    matern_spacetime_cov(
      h, u,
      sigma2_hat, beta_s_hat, beta_t_hat,
      gamma_hat, nu_hat, nug = tausq_hat
    )
  },
  h = plot_df$h,
  u = plot_df$u
)

# Convert to data frame
# plot_df <- as.data.frame(plot_df)
# 
# ggplot(plot_df, aes(x = h, y = cov, color = factor(u))) +
#   geom_line(linewidth = 1.1) +
#   scale_color_manual(values = colors) +
#   labs(
#     x = "Spatial distance h (miles)",
#     y = "Covariance C(h, u)",
#     color = "Time lag u (months)",
#     title = "Estimated Gneiting-Matérn Space–Time Covariance"
#   ) +
#   theme_minimal(base_size = 14)
```


# Space-time Kriging

```{r}
# ============================================================
# SIMPLE KRIGING (mean = 0) IN SPACE–TIME
# ============================================================

df_test <- st_test %>% arrange(ym, station)

test_y <- df_test$resid
n_train <- N
n_test  <- length(test_y)

locs_train <- as.matrix(df[, c("x", "y")])
locs_test  <- as.matrix(df_test[, c("x", "y")])

time_vec_train <- time_vec
time_vec_test  <- as.numeric(difftime(df_test$ym, min(df$ym),
                                      units = "days")) / 30.4375

D_train       <- D_spatial
D_test_train  <- rdist.earth(locs_test, locs_train, miles = TRUE)

U_train      <- U_time
U_test_train <- abs(outer(time_vec_test, time_vec_train, "-"))

K <- build_K_spacetime(sigma2_hat, beta_s_hat, nu_hat,
                       tausq_hat, beta_t_hat, gamma_hat,
                       D_train, U_train)

K_inv <- solve(K)

k <- build_K_spacetime(sigma2_hat, beta_s_hat, nu_hat,
                       nug = 0,
                       beta_t_hat, gamma_hat,
                       D_test_train, U_test_train)

known_mean <- 0
train_Z    <- y
test_Z     <- test_y

pred_test <- k %*% (K_inv %*% (train_Z - known_mean)) + known_mean

c0 <- rep(sigma2_hat + tausq_hat, n_test)

tmp      <- k %*% K_inv
var_test <- c0 - rowSums(tmp * k)

mse_test <- mean((as.vector(pred_test) - test_Z)^2, na.rm = TRUE)
cat("\nTest MSE on held-out months (residual process):", mse_test, "\n")

# ============================================================
# PREDICTIONS for next FIVE months with kriging variances
# ============================================================

# Create a regular prediction grid over the full ROI
lon_seq <- seq(-11, -5, length.out = 100)  # 100 points in longitude
lat_seq <- seq(51, 56, length.out = 100)   # 100 points in latitude
full_grid <- expand.grid(x = lon_seq, y = lat_seq)  # 10000 points

# Matrix of grid locations (needed for rdist.earth)
grid_locs <- as.matrix(full_grid[, c("x", "y")])

# ============================================================
# Krige over a spatial grid (whole Ireland) for a future month
# ============================================================

# Choose the first test month as the prediction month
target_month <- min(df_test$ym)
target_time  <- as.numeric(difftime(target_month, min(df$ym), 
                                    units = "days")) / 30.4375

# Define spatial grid (buffer ±1 degree, grid resolution = 80)
x_range <- range(coords$x)
y_range <- range(coords$y)
grid_res <- 80
grid_df <- expand.grid(
  x = seq(x_range[1] - 1, x_range[2] + 1, length.out = grid_res),
  y = seq(y_range[1] - 1, y_range[2] + 1, length.out = grid_res)
)
grid_locs <- as.matrix(grid_df[, c("x", "y")])

# Space-time distances between grid and training sites
D_grid_train <- rdist.earth(grid_locs, locs_train, miles = TRUE)
time_vec_grid <- rep(target_time, nrow(grid_df))
U_grid_train  <- abs(outer(time_vec_grid, time_vec_train, "-"))

# Build cross-covariance between grid points and training sites
k_grid <- build_K_spacetime(sigma2_hat, beta_s_hat, nu_hat,
                            nug = 0,
                            beta_t_hat, gamma_hat,
                            D_grid_train, U_grid_train)

# Simple Kriging predictions (mean = 0)
pred_grid <- as.vector(k_grid %*% (K_inv %*% train_Z))

# Kriging variance (optional)
var_grid <- sigma2_hat - rowSums(k_grid %*% K_inv * k_grid)
```


# Plot Reproduction

```{r echo=FALSE}
# Ireland map for overlay
ireland <- rnaturalearth::ne_countries(
  scale = "medium",
  country = "Ireland",
  returnclass = "sf"
)

# Global color scale based on training + prediction
z_range <- range(st_train$resid, pred_grid, na.rm = TRUE)
n_col <- 100
cols_global <- tim.colors(n_col)

# # Plot SK prediction
# quilt.plot(
#   grid_df$x, grid_df$y, pred_grid,
#   nx   = grid_res, ny = grid_res,
#   xlab = "Longitude", ylab = "Latitude",
#   main = paste0("SK prediction (residuals)\n t: ", target_month),
#   zlim = z_range,
#   col  = cols_global
# )
# par(new = TRUE)
# plot(
#   ireland$geometry,
#   border = "black",
#   lwd    = 1.4,
#   col    = NA,
#   add    = TRUE
# )

```

```{r echo=FALSE}
n_future <- 5
future_months <- max(df$ym) %m+% months(1:n_future)

pred_list <- vector("list", n_future)
var_list  <- vector("list", n_future)

# Define grid
x_range <- range(coords$x)
y_range <- range(coords$y)
grid_res <- 80
grid_df <- expand.grid(
  x = seq(x_range[1] - 1, x_range[2] + 1, length.out = grid_res),
  y = seq(y_range[1] - 1, y_range[2] + 1, length.out = grid_res)
)
grid_locs <- as.matrix(grid_df[, c("x", "y")])

for (i in seq_len(n_future)) {
  target_month <- future_months[i]
  target_time  <- as.numeric(difftime(target_month, min(df$ym), units = "days")) / 30.4375
  
  # Space-time distances between grid and training
  D_grid_train <- rdist.earth(grid_locs, locs_train, miles = TRUE)
  time_vec_grid <- rep(target_time, nrow(grid_df))
  U_grid_train  <- abs(outer(time_vec_grid, time_vec_train, "-"))
  
  # Cross-covariance
  k_grid <- build_K_spacetime(sigma2_hat, beta_s_hat, nu_hat,
                              nug = 0, beta_t_hat, gamma_hat,
                              D_grid_train, U_grid_train)
  
  # SK mean prediction
  pred_grid <- as.vector(k_grid %*% (K_inv %*% train_Z))
  
  # SK variance prediction
  var_grid <- sigma2_hat - rowSums(k_grid %*% K_inv * k_grid)
  
  pred_list[[i]] <- pred_grid
  var_list[[i]]  <- var_grid
}

names(pred_list) <- format(future_months, "%Y-%m")
names(var_list)  <- format(future_months, "%Y-%m")

```

```{r echo=FALSE}
ireland <- rnaturalearth::ne_countries(
  scale = "medium",
  country = "Ireland",
  returnclass = "sf"
)

# Color scales
n_col <- 100
cols_mean <- tim.colors(n_col)
cols_var  <- tim.colors(n_col)

z_mean <- range(unlist(pred_list), na.rm = TRUE)
z_var  <- range(unlist(var_list), na.rm = TRUE)

```

```{r echo=FALSE}
par(mfrow = c(2, 5), mar = c(4, 4, 3, 2), mgp = c(1.25, 0.5, 0))

# --- Top row: mean predictions ---
for (i in seq_len(n_future)) {
  pred_grid <- pred_list[[i]]
  
  quilt.plot(
    grid_df$x, grid_df$y, pred_grid,
    nx   = grid_res, ny = grid_res,
    xlab = "Longitude", ylab = "Latitude",
    main = paste0("Mean, t = ", names(pred_list)[i]),
    zlim = z_mean,
    col  = cols_mean
  )
  
  par(new = TRUE)
  plot(
    ireland$geometry,
    border = "black",
    lwd    = 1.4,
    col    = NA,
    add    = TRUE
  )
}

# --- Bottom row: variance predictions ---
for (i in seq_len(n_future)) {
  var_grid <- var_list[[i]]
  
  quilt.plot(
    grid_df$x, grid_df$y, var_grid,
    nx   = grid_res, ny = grid_res,
    xlab = "Longitude", ylab = "Latitude",
    main = paste0("Var, t = ", names(var_list)[i]),
    zlim = z_var,
    col  = cols_var,
  )
  
  par(new = TRUE)
  plot(
    ireland$geometry,
    border = "black",
    lwd    = 1.4,
    col    = NA,
    add    = TRUE
  )
}
```

# Open-ended questions

**What can you say about the predicted residual fields and their uncertainties? Do the spatial patterns and the increasing uncertainty over future months make sense? Why?**


For the first row of plots, the mean residual predictions are more heterogeneous to start, which reflects the initial influence of the observed wind data at each station. Areas near weather stations with high variability (like the north and southwest) show larger residuals. As time progresses though, the predictions for residuals converge toward zero. This happens because the kriging prediction is a Best Linear Unbiased Prediction. Without recent observations, the best estimate for the residual is just its expected value, because $E[\epsilon] = 0$. Temporal decay from the temporal covariance function reduces the influence of past data over time, so the spatial variability disappears over time and the mean residual map becomes less variable.

On the other hand, the predicted variance of the residuals increases over time. This is expected because the kriging variance quantifies the uncertainty of the prediction. Moving further from the observed data in time means the covariance between the prediction location and past observations decreases. Because of this, we are less certain about the residual at future times, which leads to larger predictive variances. The increasing uncertainty over time is natural for kriging with temporally decaying covariances.

Without recent data, the best estimate of the residual is its expectation, but the uncertainty grows because we know less about deviations from the mean (because we move further in time from any observed data).