---
title: "STAT 5845 Homework 7"
subtitle: "November 10, 2025"
output:
  bookdown::pdf_document2:
    latex_engine: pdflatex
    keep_tex: true 
header-includes:
  - \usepackage{titling}
  - \setlength{\droptitle}{1em} 
fontsize: 12pt
---

# Preliminaries

```{r, echo=FALSE}
# Package Import
invisible(lapply(c("dplyr", "fields", "MASS"), 
  function(pkg) suppressPackageStartupMessages(library(pkg, 
                                                character.only = TRUE))))

```

## Simulate a Gaussian random field on a 60 × 60 grid from a Matern covariance

```{r}
# Start by building a 60x60 grid of locations
x <- seq(0, 1, length.out = 60)
y <- seq(0, 1, length.out = 60)
locs <- expand.grid(x = x, y = y) %>% as.matrix()
n.loc <- nrow(locs)

# True parameters to use for simulating 
sigma2_true <- 1     # partial sill
beta_true   <- 0.05   # range parameter
nu_true     <- 2   # smoothness
tausq_true  <- 0     # nugget = 0 for simulation
mu_true     <- rep(0, n.loc)

# Build distance matrix
distance_matrix <- as.matrix(dist(locs))

# Build Matern Covaraince
covariance_matrix <- (sigma2_true / (2^(nu_true - 1) * gamma(nu_true))) *
  (distance_matrix / beta_true)^nu_true * (
    besselK(distance_matrix / beta_true, nu_true)
  )
## Handle zero distances
diag(covariance_matrix) <- sigma2_true + tausq_true

# Set random seed
set.seed(124)

# Simulate Random Field using Matern Covariance Function
Z <- mvrnorm(1, mu = mu_true, Sigma = covariance_matrix)

quilt.plot(locs[,1], locs[,2], Z, nx = 60, ny = 60)
```



## Randomly select 100 locations for training; use the remaining locations as testing.

```{r}
# Select 100 Random Points as training points
n.train <- 100
train_idx <- sample.int(n.loc, n.train) # randomly select 100 integers from 1 
                                        # to n = number of grid points

# Set the rest of the points in the grid to be used for kriging later
test_idx <- setdiff(seq_len(n.loc), train_idx)
train_locs <- locs[train_idx, , drop = FALSE]
train_Z <- Z[train_idx]
train_D <- as.matrix(dist(train_locs))
D_all_train <- fields::rdist(locs, train_locs)
```

# Exponential

```{r}
# MLE Estimation, Exponential
#-------------------- EXPONENTIAL --------------------------------------

# Note that exponential is a special case of Matern with nu = 0.5, so we can
# use the general matern MLE estimation structure, with some modifications

# Define negative log-likelihood function for MLE estimation via "optim"
nll_exponential <- function(par) {
  tausq   <- par[1]^2          # nugget variance (forced gte 0)
  sigmasq <- par[2]^2          # partial sill
  beta    <- exp(par[3])       # range parameter (gt 0)
  mu_val  <- par[4]            # constant mean
  
  # monitor parameter progress
  cat(sprintf("tau^2 = %.4f, sigma^2 = %.4f, 
              beta = %.4f, mu = %.4f\n",
              tausq, sigmasq, beta, mu_val))
  
  mu <- rep(mu_val, n.loc)
  
  dist_mat <- train_D
  # Build covariance matrix
  C <- sigmasq * exp(-dist_mat / beta)
  diag(C) <- sigmasq + tausq
  
  # Cholesky decomposition for log-likelihood
  L <- t(chol(C))
  log.det.cov <- 2 * sum(log(diag(L)))
  z.new <- forwardsolve(L, train_Z - mu)
  inner.prod <- sum(z.new^2)
  
  # Negative Gaussian log-likelihood
  nll_value <- 0.5 * n.loc * log(2*pi) + 0.5 * log.det.cov + 0.5 * inner.prod
  return(nll_value)
}
#--------------------------------------------------------------#
# Fit parameters by minimizing NLL
init <- c(0.01, 1, log(0.1), 0)
```

```{r, results='hide', message=FALSE, warning=FALSE}
fit <- optim(init, nll_exponential, control = list(trace = 0, maxit = 5000))
```

```{r}
# Verify Convergence
str(fit)

# Extract and print true vs estimated
est_par <- fit$par

results <- data.frame(
  Parameter = c("tau^2", "sigma^2", "beta (range)", "mu (mean)"),
  True = c(tausq_true, sigma2_true, beta_true, 0),
  Estimated = c(
    est_par[1]^2,
    est_par[2]^2,
    exp(est_par[3]),
    est_par[4]
  )
)

print("True vs. Estimated Parameters:")
print(results, digits = 4, row.names = FALSE)

# Plot Exponential covariance — True vs MLE 

# Distance grid for curves
dgrid <- seq(0, max(distance_matrix), length.out = 300)

# TRUE curve (simulation parameters)
cov_true <- sigma2_true * exp(-dgrid / beta_true)
cov_true[dgrid == 0] <- NA
nugget_true <- sigma2_true + tausq_true

# MLE parameters
tau2_mle   <- est_par[1]^2
sigma2_mle <- est_par[2]^2
beta_mle   <- exp(est_par[3])
nu <- 0.5

# MLE curve
cov_mle <- sigma2_mle * exp(-dgrid/beta_mle)
  
cov_mle[dgrid == 0] <- NA
nugget_mle <- sigma2_mle + tau2_mle

# Y limits 
ymax <- max(c(cov_true, cov_mle, nugget_true, nugget_mle), na.rm = TRUE)
ymin <- min(c(cov_true, cov_mle, 0), na.rm = TRUE)

plot(NA, xlim = range(dgrid), ylim = c(ymin, ymax),
     xlab = "Distance", ylab = "Covariance",
     main = "Exponential Covariance: True vs MLE")

lines(dgrid, cov_true, lwd = 2, col = "darkgreen")
lines(dgrid, cov_mle,  lwd = 2, col = "red")

# Nugget points at h = 0 
points(0, nugget_true, pch = 15, col = "darkgreen", cex = 1.2)
points(0, nugget_mle,  pch = 17, col = "red",       cex = 1.2)

legend("topright",
       legend = c("True Exponential", "MLE Exponential",
                  expression(sigma^2 + tau^2~"(True)"),
                  expression(sigma^2 + tau^2~"(MLE)")),
       lty    = c(1, 1, NA, NA),
       pch    = c(NA, NA, 15, 17),
       col    = c("darkgreen", "red", "darkgreen", "red"),
       bty    = "n")
```

```{r}
#------------------ KRIGING PHASE -----------------------------------

# Build Sigma(\hat{theta}) for training data (retrieve MLE estiamtes again)
tau2_hat   <- est_par[1]^2
sigma2_hat <- est_par[2]^2
beta_hat   <- exp(est_par[3])
mu_hat     <- est_par[4]

# Covariance matrix using exponential model
exp_cov <- function(D, sigma2, beta) {
  sigma2 * exp(-D / beta)
}

C_tt <- exp_cov(train_D, sigma2_hat, beta_hat)
diag(C_tt) <- sigma2_hat + tau2_hat + 1e-10   # ensure numerical stability

# Find Explicit inverse for kriging
Sigma_inv <- solve(C_tt)

# Get Design matrix for mean m(s) (intercept-only model)
M <- matrix(1, nrow = n.train, ncol = 1)
beta_hat_gls <- solve(t(M) %*% Sigma_inv %*% M) %*%
  (t(M) %*% Sigma_inv %*% matrix(train_Z, ncol = 1))

# Cross-covariances K(\hat{theta}) and predictions at all grid location
K_st <- exp_cov(D_all_train, sigma2_hat, beta_hat)
K_st[D_all_train == 0] <- sigma2_hat

m0 <- matrix(1, nrow = n.loc, ncol = 1)
resid <- matrix(train_Z, ncol = 1) - M %*% beta_hat_gls
pred_all <- as.vector(m0 %*% beta_hat_gls + K_st %*% (Sigma_inv %*% resid))

# Kriging variances
C_ss_diag <- rep(sigma2_hat + tau2_hat, n.loc)
tmp <- K_st %*% Sigma_inv
var_all <- C_ss_diag - rowSums(tmp * K_st)

# Test MSE on held-out points
mse_test <- mean((pred_all[test_idx] - Z[test_idx])^2)
```

``` {r fig.width=8.5, fig.height=2.75}
# Plots (Truth | Training | Prediction + MSE)
par(mfrow = c(1,3))
quilt.plot(locs[,1], locs[,2], Z, nx = length(x), 
           ny = length(y), main = "Truth")
quilt.plot(train_locs[,1], train_locs[,2], train_Z, nx = length(x), 
           ny = length(y), main = "Training")
quilt.plot(locs[,1], locs[,2], pred_all, nx = length(x), ny = length(y),
           main = paste0("Kriging (Exponential)\nTest MSE = ", 
                         signif(mse_test, 4)))
points(train_locs[,1], train_locs[,2], pch = 16, cex = 0.3)

print(paste("Exponential Covariance MSE:", round(mse_test,4)))
```

---

# Gaussian

```{r}
# MLE Estimation, Gaussian
#-------------------- GAUSSIAN -----------------------------------------
# Define negative log-likelihood function for MLE estimation via "optim"
nll_gaussian <- function(par) {
  tausq   <- par[1]^2          # nugget variance (forced gte 0)
  sigmasq <- par[2]^2          # partial sill
  beta    <- exp(par[3])       # range parameter (gt 0)
  mu_val  <- par[3]            # constant mean
  
  # monitor parameter progress
  cat(sprintf("tau^2 = %.4f, sigma^2 = %.4f, 
              beta = %.4f, mu = %.4f\n",
              tausq, sigmasq, beta, nu, mu_val))
  
  mu <- rep(mu_val, n.loc)
  
  dist_mat <- train_D
  # Build covariance matrix
  C <- sigmasq * exp(-(dist_mat / beta)^2)
  diag(C) <- sigmasq + tausq
  
  # Cholesky decomposition for log-likelihood
  L <- t(chol(C))
  log.det.cov <- 2 * sum(log(diag(L)))
  z.new <- forwardsolve(L, train_Z - mu)
  inner.prod <- sum(z.new^2)
  
  # Negative Gaussian log-likelihood
  nll_value <- 0.5 * n.loc * log(2*pi) + 0.5 * log.det.cov + 0.5 * inner.prod
  return(nll_value)
}
#--------------------------------------------------------------#
# Fit parameters by minimizing NLL
init <- c(0.01, 1, log(0.1), 0)
```

```{r, results='hide', message=FALSE, warning=FALSE}
fit <- optim(init, nll_gaussian, control = list(trace = 0, maxit = 5000))
```

```{r}
# Verify Convergence
str(fit)

# Extract and print true vs estimated
est_par <- fit$par

results <- data.frame(
  Parameter = c("tau^2", "sigma^2", "beta (range)", "mu (mean)"),
  True = c(tausq_true, sigma2_true, beta_true, 0),
  Estimated = c(
    est_par[1]^2,
    est_par[2]^2,
    exp(est_par[3]),
    est_par[4]
  )
)

print("True vs. Estimated Parameters:")
print(results, digits = 4, row.names = FALSE)

# Plot Gaussian covariance — True vs MLE 

# Distance grid for curves
dgrid <- seq(0, max(distance_matrix), length.out = 300)

# TRUE curve (simulation parameters)
cov_true <- sigma2_true * exp(-(dgrid / beta_true)**2)
cov_true[dgrid == 0] <- NA
nugget_true <- sigma2_true + tausq_true

# MLE parameters
tau2_mle   <- est_par[1]^2
sigma2_mle <- est_par[2]^2
beta_mle   <- exp(est_par[3])

# MLE curve
cov_mle <- sigma2_mle * exp(-(dgrid/beta_mle)**2)
cov_mle[dgrid == 0] <- NA
nugget_mle <- sigma2_mle + tau2_mle

# Y limits 
ymax <- max(c(cov_true, cov_mle, nugget_true, nugget_mle), na.rm = TRUE)
ymin <- min(c(cov_true, cov_mle, 0), na.rm = TRUE)

plot(NA, xlim = range(dgrid), ylim = c(ymin, ymax),
     xlab = "Distance", ylab = "Covariance",
     main = "Gaussian Covariance: True vs MLE")

lines(dgrid, cov_true, lwd = 2, col = "darkgreen")
lines(dgrid, cov_mle,  lwd = 2, col = "red")

# Nugget points at h = 0 
points(0, nugget_true, pch = 15, col = "darkgreen", cex = 1.2)
points(0, nugget_mle,  pch = 17, col = "red",       cex = 1.2)

legend("topright",
       legend = c("True Gaussian", "MLE Gaussian",
                  expression(sigma^2 + tau^2~"(True)"),
                  expression(sigma^2 + tau^2~"(MLE)")),
       lty    = c(1, 1, NA, NA),
       pch    = c(NA, NA, 15, 17),
       col    = c("darkgreen", "red", "darkgreen", "red"),
       bty    = "n")
```

```{r}
#------------------ KRIGING PHASE -----------------------------------

# Extract MLE parameter estimates
tau2_hat   <- est_par[1]^2
sigma2_hat <- est_par[2]^2
beta_hat   <- exp(est_par[3])
mu_hat     <- est_par[4]

# Define Gaussian covariance function
gaussian_cov <- function(D, sigma2, beta) {
  sigma2 * exp(-(D / beta)^2)
}

# Build covariance matrix for training data
C_tt <- gaussian_cov(train_D, sigma2_hat, beta_hat)
diag(C_tt) <- sigma2_hat + tau2_hat + 1e-10   # ensure positive definiteness

# Compute inverse for kriging
Sigma_inv <- solve(C_tt)

# Design matrix for mean m(s) (intercept-only model)
M <- matrix(1, nrow = n.train, ncol = 1)
beta_hat_gls <- solve(t(M) %*% Sigma_inv %*% M) %*%
  (t(M) %*% Sigma_inv %*% matrix(train_Z, ncol = 1))

# Cross-covariances k(\hat{theta}) and predictions at all grid location
K_st <- gaussian_cov(D_all_train, sigma2_hat, beta_hat)
K_st[D_all_train == 0] <- sigma2_hat

m0 <- matrix(1, nrow = n.loc, ncol = 1)
resid <- matrix(train_Z, ncol = 1) - M %*% beta_hat_gls
pred_all <- as.vector(m0 %*% beta_hat_gls + K_st %*% (Sigma_inv %*% resid))

# Kriging variances
C_ss_diag <- rep(sigma2_hat + tau2_hat, n.loc)
tmp <- K_st %*% Sigma_inv
var_all <- C_ss_diag - rowSums(tmp * K_st)

# Test MSE on held-out points
mse_test <- mean((pred_all[test_idx] - Z[test_idx])^2)
```

``` {r fig.width=8.5, fig.height=2.75}
# Visualization
par(mfrow = c(1,3))
quilt.plot(locs[,1], locs[,2], Z, nx = length(x), ny = length(y),
           main = "Truth")
quilt.plot(train_locs[,1], train_locs[,2], train_Z, nx = length(x), 
           ny = length(y), main = "Training")
quilt.plot(locs[,1], locs[,2], pred_all, nx = length(x), ny = length(y),
           main = paste0("Kriging (Gaussian)\nTest MSE = ", 
                         signif(mse_test, 4)))
points(train_locs[,1], train_locs[,2], pch = 16, cex = 0.3)

print(paste("Gaussian Covariance MSE:", round(mse_test,4)))
```

---

# Matern

```{r}
# MLE Estimation, Matern
#-------------------- MATERN -------------------------------------------
# Define negative log-likelihood function for MLE estimation via "optim"
nll_matern <- function(par) {
  tausq   <- par[1]^2          # nugget variance (forced gte 0)
  sigmasq <- par[2]^2          # partial sill
  beta    <- exp(par[3])       # range parameter (gt 0)
  nu      <- par[4]            # smoothness
  mu_val  <- par[5]            # constant mean
  
  # monitor parameter progress
  cat(sprintf("tau^2 = %.4f, sigma^2 = %.4f, 
              beta = %.4f, nu = %.4f, mu = %.4f\n",
              tausq, sigmasq, beta, nu, mu_val))
  
  mu <- rep(mu_val, n.loc)
  
  dist_mat <- train_D
  # Build covariance matrix
  C <- (sigmasq / (2^(nu - 1) * gamma(nu))) *
      (dist_mat / beta)^nu * besselK(dist_mat / beta, nu)
  diag(C) <- sigmasq + tausq
  
  # Cholesky decomposition for log-likelihood
  L <- t(chol(C))
  log.det.cov <- 2 * sum(log(diag(L)))
  z.new <- forwardsolve(L, train_Z - mu)
  inner.prod <- sum(z.new^2)
  
  # Negative Gaussian log-likelihood
  nll_value <- 0.5 * n.loc * log(2*pi) + 0.5 * log.det.cov + 0.5 * inner.prod
  return(nll_value)
}
#--------------------------------------------------------------#

# Fit parameters by minimizing NLL
init <- c(0.1, 1, log(0.1), 0.8, 0)
```

```{r, results='hide', message=FALSE, warning=FALSE}
fit <- optim(init, nll_matern, control = list(trace = 0, maxit = 5000))
```

```{r}
# Verify Convergence
str(fit)

# Extract and print true vs estimated
est_par <- fit$par

results <- data.frame(
  Parameter = c("tau^2", "sigma^2", "beta (range)", 
                "nu (smoothness)", "mu (mean)"),
  True = c(tausq_true, sigma2_true, beta_true, nu_true, 0),
  Estimated = c(
    est_par[1]^2,
    est_par[2]^2,
    exp(est_par[3]),
    est_par[4],
    est_par[5]
  )
)

print("True vs. Estimated Parameters:")
print(results, digits = 4, row.names = FALSE)

# Plot Matern covariance — True vs MLE 

# Distance grid for curves
dgrid <- seq(0, max(distance_matrix), length.out = 300)

# TRUE curve (simulation parameters)
cov_true <- (sigma2_true / (2^(nu_true - 1) * gamma(nu_true))) *
  (dgrid / beta_true)^nu_true * besselK(dgrid / beta_true, nu_true)
cov_true[dgrid == 0] <- NA
nugget_true <- sigma2_true + tausq_true

# MLE parameters
tau2_mle   <- est_par[1]^2
sigma2_mle <- est_par[2]^2
beta_mle   <- exp(est_par[3])
nu_mle     <- est_par[4]

# MLE curve
cov_mle <- (sigma2_mle / (2^(nu_mle - 1) * gamma(nu_mle))) *
  (dgrid / beta_mle)^nu_mle * besselK(dgrid / beta_mle, nu_mle)
cov_mle[dgrid == 0] <- NA
nugget_mle <- sigma2_mle + tau2_mle

# Y limits 
ymax <- max(c(cov_true, cov_mle, nugget_true, nugget_mle), na.rm = TRUE)
ymin <- min(c(cov_true, cov_mle, 0), na.rm = TRUE)

plot(NA, xlim = range(dgrid), ylim = c(ymin, ymax),
     xlab = "Distance", ylab = "Covariance",
     main = "Matern Covariance: True vs MLE")

lines(dgrid, cov_true, lwd = 2, col = "darkgreen")
lines(dgrid, cov_mle,  lwd = 2, col = "red")

# Nugget points at h = 0 
points(0, nugget_true, pch = 15, col = "darkgreen", cex = 1.2)
points(0, nugget_mle,  pch = 17, col = "red",       cex = 1.2)

legend("topright",
       legend = c("True Matern", "MLE Matern",
                  expression(sigma^2 + tau^2~"(True)"),
                  expression(sigma^2 + tau^2~"(MLE)")),
       lty    = c(1, 1, NA, NA),
       pch    = c(NA, NA, 15, 17),
       col    = c("darkgreen", "red", "darkgreen", "red"),
       bty    = "n")
```

```{r}
#------------------ KRIGING PHASE -----------------------------------

# Extract MLE parameter estimates
tau2_hat   <- est_par[1]^2
sigma2_hat <- est_par[2]^2
beta_hat   <- exp(est_par[3])
nu_hat     <- est_par[4]
mu_hat     <- est_par[5]

# Define Matern covariance function
matern_cov <- function(D, sigma2, beta, nu) {
  # Handle zero distances to avoid NaN from BesselK
  D[D == 0] <- 1e-10
  cov_val <- (sigma2 / (2^(nu - 1) * gamma(nu))) *
    (D / beta)^nu * besselK(D / beta, nu)
  return(cov_val)
}

# Build covariance matrix for training data
C_tt <- matern_cov(train_D, sigma2_hat, beta_hat, nu_hat)
diag(C_tt) <- sigma2_hat + tau2_hat + 1e-10   # ensure positive definiteness

# Compute inverse for kriging
Sigma_inv <- solve(C_tt)

# Design matrix for mean m(s) (intercept-only model)
M <- matrix(1, nrow = n.train, ncol = 1)
beta_hat_gls <- solve(t(M) %*% Sigma_inv %*% M) %*%
  (t(M) %*% Sigma_inv %*% matrix(train_Z, ncol = 1))

# Cross-covariances k(\hat{theta}) and predictions at all grid location
K_st <- matern_cov(D_all_train, sigma2_hat, beta_hat, nu_hat)
K_st[D_all_train == 0] <- sigma2_hat

m0 <- matrix(1, nrow = n.loc, ncol = 1)
resid <- matrix(train_Z, ncol = 1) - M %*% beta_hat_gls
pred_all <- as.vector(m0 %*% beta_hat_gls + K_st %*% (Sigma_inv %*% resid))

# Kriging variances
C_ss_diag <- rep(sigma2_hat + tau2_hat, n.loc)
tmp <- K_st %*% Sigma_inv
var_all <- C_ss_diag - rowSums(tmp * K_st)

# Test MSE on held-out points
mse_test <- mean((pred_all[test_idx] - Z[test_idx])^2)
```

``` {r fig.width=8.5, fig.height=2.75}
# Visualization
par(mfrow = c(1,3))
quilt.plot(locs[,1], locs[,2], Z, nx = length(x), ny = length(y),
           main = "Truth")
quilt.plot(train_locs[,1], train_locs[,2], train_Z, nx = length(x), 
           ny = length(y), main = "Training")
quilt.plot(locs[,1], locs[,2], pred_all, nx = length(x), ny = length(y),
           main = paste0("Kriging (Matern, nu = ", round(nu_hat, 2), 
                         ")\nTest MSE = ", signif(mse_test, 4)))
points(train_locs[,1], train_locs[,2], pch = 16, cex = 0.3)

print(paste("Matern Covariance MSE:", round(mse_test, 4)))
```

# Concluding Question: What do you observe about the kriging predictions under the Exponential, Gaussian, and Matern models? Which performs best based on the MSE, and why?

\textbf{Based on the MLE estimates obtained here, the predictions made under the MLE-estimated Matern covariance function had the lowest MSE, followed closely by the MLE-approximated exponential covariance function. The Gaussian covariance MLE estimates did not prove to be very effective for this example, having a much higher MSE. Graphically, we observe an overly-smoothed prediction for the Gaussian MLE covariance function. While this covariance function had the worst predictions, we do see that the covariance function provided reasonable sensitivity for interpolation using kriging, though it missed some of the finer details of the trend.}

\textbf{The MLE-estimated exponential covariance function also had troubling predictions, missing finer details while capturing the general trends reasonably. Relatively, the MLE-esimated Matern covariance function did a bit of a better job at capturing some finer trends, but the fit had imperfections much like the other models.}

\textbf{Overall, we find that MLE estimation of covariance functions followed by kriging allows us to interpolate surfaces that make the ``magic'' of spatial prediction possible---a very valuable tool. However, we see that our predictions are limited by the effectiveness of maximum likelihood estimation for our covariance model parameters and the count of training data points given to interpolate with.}

\emph{Special credit to Dr. Salvana for providing much of the relevant code for this assignment.}