---
title: "STAT 5845 Homework 7"
subtitle: "November 10, 2025"
output:
  bookdown::pdf_document2:
    latex_engine: pdflatex
    keep_tex: true 
header-includes:
  - \usepackage{titling}
  - \setlength{\droptitle}{1em} 
fontsize: 12pt
---

# Preliminaries

```{r, echo=FALSE}
# Package Import
invisible(lapply(c("dplyr", "fields", "MASS"), 
  function(pkg) suppressPackageStartupMessages(library(pkg, 
                                                character.only = TRUE))))

```

## Simulate a Gaussian random field on a 60 × 60 grid from a Matérn covariance

```{r}
# Start by building a 60x60 grid of locations
x <- seq(0, 1, length.out = 60)
y <- seq(0, 1, length.out = 60)
locs <- expand.grid(x = x, y = y) %>% as.matrix()
n.loc <- nrow(locs)

# True parameters to use for simulating 
sigma2_true <- 1     # partial sill
beta_true   <- 0.2   # range parameter
nu_true     <- 1.0   # smoothness
tausq_true  <- 0     # nugget = 0 for simulation
mu_true     <- rep(0, n.loc)

# Build distance matrix
distance_matrix <- as.matrix(dist(locs))

# Build Matern Covaraince
covariance_matrix <- (sigma2_true / (2^(nu_true - 1) * gamma(nu_true))) *
  (distance_matrix / beta_true)^nu_true * besselK(distance_matrix / beta_true, nu_true)
## Handle zero distances
diag(covariance_matrix) <- sigma2_true + tausq_true

# Set random seed
set.seed(124)

# Simulate Random Field using Matern Covariance Function
Z <- mvrnorm(1, mu = mu_true, Sigma = covariance_matrix)

dev.new(width = 6, height = 6)
quilt.plot(locs[,1], locs[,2], Z, nx = 60, ny = 60)
```

## Randomly select 100 locations for training; use the remaining locations as testing.

```{r}
# Select 100 Random Points as training points
n.train <- 100
train_idx <- sample.int(n.loc, n.train) # randomly select 100 integers from 1 
                                        # to n = number of grid points

# Set the rest of the points in the grid to be used for kriging later
test_idx <- setdiff(seq_len(n.loc), train_idx)
train_locs <- locs[train_idx, , drop = FALSE]
train_Z <- Z[train_idx]
train_D <- as.matrix(dist(train_locs))
D_all_train <- fields::rdist(locs, train_locs)
```

## Estimate parameters by MLE on the training data under three models: Exponential, Gaussian, and Matérn

```{r}
# MLE Estimation, Exponential
#-------------------- EXPONENTIAL --------------------------------------
# Define negative log-likelihood function for MLE estimation via "optim"
nll_exponential <- function(par) {
  tausq   <- par[1]^2          # nugget variance (forced ≥ 0)
  sigmasq <- par[2]^2          # partial sill
  beta    <- exp(par[3])       # range parameter (> 0)
  nu      <- par[4]            # smoothness
  mu_val  <- par[5]            # constant mean
  
  # monitor parameter progress
  cat(sprintf("tau^2 = %.4f, sigma^2 = %.4f, 
              beta = %.4f, nu = %.4f, mu = %.4f\n",
              tausq, sigmasq, beta, nu, mu_val))
  
  mu <- rep(mu_val, n.loc)
  
  dist_mat <- distance_matrix
  # Build covariance matrix
  C <- sigmasq * exp(-dist_mat / beta)
  diag(C) <- sigmasq + tausq
  
  # Cholesky decomposition for log-likelihood
  L <- t(chol(C))
  log.det.cov <- 2 * sum(log(diag(L)))
  z.new <- forwardsolve(L, Z - mu)
  inner.prod <- sum(z.new^2)
  
  # Negative Gaussian log-likelihood
  nll_value <- 0.5 * n.loc * log(2 * pi) + 0.5 * log.det.cov + 0.5 * inner.prod
  return(nll_value)
}
#--------------------------------------------------------------#

```

```{r}
# MLE Estimation, Gaussian
#-------------------- GAUSSIAN -----------------------------------------
# Define negative log-likelihood function for MLE estimation via "optim"
nll_gaussian <- function(par) {
  tausq   <- par[1]^2          # nugget variance (forced ≥ 0)
  sigmasq <- par[2]^2          # partial sill
  beta    <- exp(par[3])       # range parameter (> 0)
  nu      <- par[4]            # smoothness
  mu_val  <- par[5]            # constant mean
  
  # monitor parameter progress
  cat(sprintf("tau^2 = %.4f, sigma^2 = %.4f, 
              beta = %.4f, nu = %.4f, mu = %.4f\n",
              tausq, sigmasq, beta, nu, mu_val))
  
  mu <- rep(mu_val, n.loc)
  
  dist_mat <- distance_matrix
  # Build covariance matrix
  C <- sigmasq * exp(-(dist_mat / beta)^2)
  diag(C) <- sigmasq + tausq
  
  # Cholesky decomposition for log-likelihood
  L <- t(chol(C))
  log.det.cov <- 2 * sum(log(diag(L)))
  z.new <- forwardsolve(L, Z - mu)
  inner.prod <- sum(z.new^2)
  
  # Negative Gaussian log-likelihood
  nll_value <- 0.5 * n.loc * log(2 * pi) + 0.5 * log.det.cov + 0.5 * inner.prod
  return(nll_value)
}
#--------------------------------------------------------------#

```

```{r}
# MLE Estimation, Matern
#-------------------- MATERN -------------------------------------------
# Define negative log-likelihood function for MLE estimation via "optim"
nll_matern <- function(par) {
  tausq   <- par[1]^2          # nugget variance (forced ≥ 0)
  sigmasq <- par[2]^2          # partial sill
  beta    <- exp(par[3])       # range parameter (> 0)
  nu      <- par[4]            # smoothness
  mu_val  <- par[5]            # constant mean
  
  # monitor parameter progress
  cat(sprintf("tau^2 = %.4f, sigma^2 = %.4f, 
              beta = %.4f, nu = %.4f, mu = %.4f\n",
              tausq, sigmasq, beta, nu, mu_val))
  
  mu <- rep(mu_val, n.loc)
  
  dist_mat <- distance_matrix
  # Build covariance matrix
  C <- (sigmasq / (2^(nu - 1) * gamma(nu))) *
      (dist_mat / beta)^nu * besselK(dist_mat / beta, nu)
  diag(C) <- sigmasq + tausq
  
  # Cholesky decomposition for log-likelihood
  L <- t(chol(C))
  log.det.cov <- 2 * sum(log(diag(L)))
  z.new <- forwardsolve(L, Z - mu)
  inner.prod <- sum(z.new^2)
  
  # Negative Gaussian log-likelihood
  nll_value <- 0.5 * n.loc * log(2 * pi) + 0.5 * log.det.cov + 0.5 * inner.prod
  return(nll_value)
}
#--------------------------------------------------------------#

# Fit parameters by minimizing NLL
init <- c(0.1, 1, log(0.3), 0.8, 0)
fit <- optim(init, nll_matern, control = list(trace = 3, maxit = 5000))
#fit <- optim(par = init, fn = nll, method = "BFGS", control = list(maxit = 5000))

# Verify Convergence
str(fit)

# Extract and print true vs estimated
est_par <- fit$par

results <- data.frame(
  Parameter = c("tau^2", "sigma^2", "beta (range)", "nu (smoothness)", "mu (mean)"),
  True = c(tausq_true, sigma2_true, beta_true, nu_true, 0),
  Estimated = c(
    est_par[1]^2,
    est_par[2]^2,
    exp(est_par[3]),
    est_par[4],
    est_par[5]
  )
)

print("True vs. Estimated Parameters:")
print(results, digits = 4, row.names = FALSE)

# Plot Matérn covariance — True vs MLE 

# Distance grid for curves
dgrid <- seq(0, max(distance_matrix), length.out = 300)

# TRUE curve (simulation parameters)
cov_true <- (sigma2_true / (2^(nu_true - 1) * gamma(nu_true))) *
  (dgrid / beta_true)^nu_true * besselK(dgrid / beta_true, nu_true)
cov_true[dgrid == 0] <- NA
nugget_true <- sigma2_true + tausq_true

# MLE parameters
tau2_mle   <- est_par[1]^2
sigma2_mle <- est_par[2]^2
beta_mle   <- exp(est_par[3])
nu_mle     <- est_par[4]

# MLE curve
cov_mle <- (sigma2_mle / (2^(nu_mle - 1) * gamma(nu_mle))) *
  (dgrid / beta_mle)^nu_mle * besselK(dgrid / beta_mle, nu_mle)
cov_mle[dgrid == 0] <- NA
nugget_mle <- sigma2_mle + tau2_mle

# Y limits 
ymax <- max(c(cov_true, cov_mle, nugget_true, nugget_mle), na.rm = TRUE)
ymin <- min(c(cov_true, cov_mle, 0), na.rm = TRUE)

plot(NA, xlim = range(dgrid), ylim = c(ymin, ymax),
     xlab = "Distance", ylab = "Covariance",
     main = "Matérn Covariance: True vs MLE")

lines(dgrid, cov_true, lwd = 2, col = "darkgreen")
lines(dgrid, cov_mle,  lwd = 2, col = "red")

# Nugget points at h = 0 
points(0, nugget_true, pch = 15, col = "darkgreen", cex = 1.2)
points(0, nugget_mle,  pch = 17, col = "red",       cex = 1.2)

legend("topright",
       legend = c("True Matérn", "MLE Matérn",
                  expression(sigma^2 + tau^2~"(True)"),
                  expression(sigma^2 + tau^2~"(MLE)")),
       lty    = c(1, 1, NA, NA),
       pch    = c(NA, NA, 15, 17),
       col    = c("darkgreen", "red", "darkgreen", "red"),
       bty    = "n")
```

## Perform kriging to predict at all grid locations for each fitted model

# Task 1:

```{r}

```

# Task 2:

```{r}

```

# Task 3: 

```{r}

```

# Concluding Question: What do you observe about the kriging predictions under the Exponential, Gaussian, and Matérn models? Which performs best based on the MSE, and why?

\textbf{ANSWER}

\emph{Special credit to Dr. Salvaña for providing much of the relevant code for this assignment.}